Researchers' perceptions of malaria eradication: findings from a mixed-methods analysis of a large survey

### 2018-02-13 paper.Rmd
adjusting to jacqulines comments

---
title: "Researchers' perceptions of malaria eradication: findings from a mixed-methods analysis of a large survey"
bibliography: library.bib
output: 
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: preamble.sty
---

@@

# Basic knitr options
opts_chunk$set(comment = NA, 
               echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE,
               fig.align = 'center')
options(xtable.comment = FALSE)

@@

Many studies have shown value in expert elicitation as a means to reduce uncertainty and inform decision making [@Morgan_2014]. But experts can sometimes be mistaken, a fact implicitly recognized when patients request a "second opinion" from another doctor when facing uncertainty. Additionally, relying on only a small panel of experts for a problem as complex as malaria exposes one to more bias, just as an overly concentrated investment portfolio exposes one to more risk. As Sir Francis Galton demonstrated in his famous study in which he showed that the crowds' aggregated estimates of cow's weight formed a quasi-normal distribution centered around the true weight [@Wallis2014] [@Galton1907Voxpopuli], averaging the perceptions of many can be more accurate than taking the opinion of any single expert, since the biases of diverse viewpoints can be complementary and symbiotic.

@@

The objective of this study is to help guide the optimal distribution of health resources by gauging (expert) opinion about, estimating the likelihood of, and quantifying the time frame to malaria eradication through a systemic survey of malaria research professionals from a wide array of academic disciplines.

@@

1. Area of expertise.
2. Perceived probability (%) of malaria eradication in 10, 20, 30, 40, and 50 years.
3. Free choice perceived number of years until malaria eradication.
4. Ranking of the WHO's "health system building blocks" [@manya2017] in regards to attention needed in order to achieve eradication.

The purpose of final point was to introduce a prescriptive element to the researchers' responses (ie, identify potential areas of consensus in regards to the path towards eradication).

The survey was intentionally as short as possible, so as to appeal to time-pressed participants. However, supplementary data on researchers is of value for the assessing of selection bias and determinants of perception. Accordingly, we estimated the likelihood of a participant being of one of two genders (male or female) based on first name, and the likelihood of a participant being of one of five ethnicities (White, Black, Hispanic, Asian, Other) based on the provided last name. For gender estimations we used data from the North Atlantic Population Project, and U.S. government [@Lincoln]; for ethnicity, Khanna et al's bayesian prediction algorithm for the estimation of racial categories [@wru2015] [@Imai_2016] was employed.

The searching and retrieval of information from the PubMed database was carried out using the RISmed package [@Kovalchik2015].

@@

This research followed the principles of the Declaration of Helsinki. The study's protocol was approved by the Scientific Committee of the Barcelona Institute for Global Health.

@@

@@

Respondents were qualitatively different from non-respondents. Importantly, the average number of total author-specific citations was `r round(mean(df$n_citations_total[df$responded]), digits = 2)` among respondents, but `r round(mean(df$n_citations_total[!df$responded]), digits = 2)` among non-respondents. This suggests either a tendency for more senior or impactful researchers not to respond. When examining the number of average citations per article, the difference between respondents remained: `r round(mean(df$n_citations_average[df$responded]), digits = 2)` among respondents, and `r round(mean(df$n_citations_average[!df$responded]), digits = 2)` among non-respondents, highlighting the greater impact of non-respondents relative to respondents. Males responded at a greater rate (`r round(mean(df$responded[df$sex == 'male'], na.rm = T) * 100, digits = 2)`) than females (`r round(mean(df$responded[df$sex == 'female'], na.rm = T) * 100, digits = 2)`) and those with Hispanic last names responded at a greater rate than those with last names of other ethnicities (see table).

@@

\subsection*{De-biasing sample selection}

In order to de-bias sample selection, we employed Heckman's two-step correction method as well a more simple binomial logistic regression model to estimate the likelihood of response as a function of sex and (binned) number of citations. To the extent that the two methods presented nearly identical value, we present here the results of latter. Having estimated the odds of survey participation, we then used the inverse of the selection model's predictions as weights in a simple linear model to adjust our estimates. We run weighted model to estimate the likelihood of eradication at 10, 20, 30, 40, and 50 years (employing a separate model for each time period). Figure 3 shows both the aggregated perceived likelihood of eradication over time by number of publications (panel A) as well as the difference in the aggregate perceived likelihood of eradication for the entire sample before and after adjusting for sample selection (panel B).

@@

}
simple <- gather(simple, key, value, adjusted:unadjusted)
g2 <- ggplot(data = simple,
       aes(x = years,
           y = value,
           color = key,
           lty = key)) +
    geom_line(alpha = 0.8) +
  geom_point(alpha = 0.6) +
  theme_paper(base_size = 13) +
  labs(x = 'Years',
       y = 'Likelihood',
       subtitle = 'Panel B') +
  scale_linetype_manual(name = '', values = 1:2) +
  scale_color_manual(name = '', values = c('blue', 'darkorange'))
x <- df %>%
  group_by(citations = bin_citations) %>%
  summarise(years_10 = mean(years_10_numeric, na.rm = TRUE),
            years_20 = mean(years_20_numeric, na.rm = TRUE),
            years_30 = mean(years_30_numeric, na.rm = TRUE),
            years_40 = mean(years_40_numeric, na.rm = TRUE),
            years_50 = mean(years_50_numeric, na.rm = TRUE),
            n = length(unique(email))) %>%
  ungroup %>%
  gather(key, value, years_10:years_50) %>%
  filter(!is.na(citations))

cols <- colorRampPalette(c('red', 'orange', 'blue') #brewer.pal(n = 9, 'Spectral')
                         )(length(unique(x$citations)))
g1 <- ggplot(data = x %>% filter(!is.na(citations)),
       aes(x = as.numeric(gsub('years_', '', key)),
           y = value,
           color = citations,
           group = citations)) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.6) +
  theme_paper(base_size = 13) +
  scale_color_manual(name = 'N° citations',
                     values = cols) +
  labs(x = 'Years',
       y = 'Likelihood',
       subtitle = 'Panel A')
```

@@

2. *Systematic challenges*: comments pertaining to political, social, environmental or logistical issues related to eradication.

3. *Complexity*: comments which of focus on the multi-dimensional components of eradication.

@@

Others echoed the sentiment, with many comments focusing on the need of strong surveillance and treatment delivery systems. Many commenters focused on other reasons for stagnating progress, such as "weak or failing health systems...due to political unwillingness or conflict".  Comments coded as "systemic" tended to be descriptive and more pessimistic than others. Many noted that malaria is a "disease of poverty", with "social injustice" as the root cause. Some made the sequential argument that "eradication of poverty" must preceed disease eradication. Along the same lines, one commenter wrote:

@@

Another noted that the survey "left off the list the most important factor - economic development". Many echoed the sentiment, stating that "with no economic development, you cannot have eradication" and that poverty is the "cause" of malaria. As with the "systemic challenges" of eradication, comments which were coded as poverty tended to be descriptive rather than prescriptive.

@@

Comments questioning the utility of eradication as a concept or goal tended to be skeptical of the its feasibility. Largely, they were prescriptive, advocating for a re-framing of the conversation so that the focus was not on an "arbitrary" goal like eradication, but rather on scaling up control and making region-specific progress.

@@

Our survey reveals a notable gap between the public discourse on malaria eradication, and those views held (largely privately) by malaria researchers. Approximately three-quarters of respondents believe that malaria will not be eradicated in the next 50 years; in other words, assuming our pool of respondents is of typical post-PhD age, most believe they will not live to see eradication. When adjusted for selection bias, the perceived likelihood of eradication 50 years from now remains similarly low, but estimates for shorter-term eradication are even lower.

If we take the pooled opinions of researchers, as revealed through this survey, as the best approximation of truth available, then the disconnect between public discourse on malaria eradication and private opinion is troubling. This has important implications for spending in public health, since the attention and resources of funders, researchers and government agencies are directed to those areas where a result is expected.

@@

There are several reasons to be skeptical that the combined "wisdom of crowds" of malaria researchers is a reliable indicator of truth. First, academic researchers are specialists - their narrow, field-specific view of eradication's feasibility is arguably less reliable than the more "global" views of large institutions. Though crowds have been found to be more "wise" than individuals in many cases, the application of a wisdom of crowds approach is not suitable to all classes of problems [@Mannes_2014]. Crowds can be susceptible to social biases [@Lorenz_2011] (although this survey's anonimity largely protects against this issue), and other biases may come in to play, especially given that our study was of a crowd of "specialists", rather than the population as a whole.

@@

Our survey attempted to quantify the likelihood of and time frame to malaria eradication by gauging the collective opinion of malaria researchers. Our survey shows that the opinions of malaria researchers appear to diverge from those of malaria programme financiers in regards to the feasibility and timeframe of eradication, with researchers expressing more pessism than the institutions and individuals which finance and oversee global eradication efforts. The causes for pessimism are diverse, but common themes were the need for innovation, systemic challenges, and the complexity of the disease and its transmissio.

@@

We have strived to present the results of our study without insinuation regarding the "true" feasibility and timeframe of eradication, and only time will tell whether the collective "wisdom" of researchers was worth adhering to or not. If the research community is unduly pessimistic, efforts should be made to convince them that eradication is more feasible than they realize (so that resources are directed accordingly). If, on the other hand, collective wisdom is a better forecasting agent than the stated objectives of global health leaders and institutions, then it is important to recognize the sad truth that we are likely further from eradication than we would like to be. In this case, though the discourse on eradication can be motivating and inspiring, resources should be directed in a manner that reflects an intended outcome's likelihood of occurrence.

>>>>

---
title: "Researchers' perceptions of malaria eradication: findings from a mixed-methods analysis of a large survey"
bibliography: library.bib
output: 
  pdf_document:
    latex_engine: xelatex
    includes:
      in_header: preamble.sty
csl: elsevier-with-titles.csl
---

@@

# Basic knitr options
opts_chunk$set(comment = NA, 
               echo = FALSE, 
               warning = FALSE, 
               message = FALSE, 
               error = TRUE, 
               cache = FALSE,
               fig.align = 'center',
               fig.pos = 'h')
options(xtable.comment = FALSE)

@@

Many studies have shown value in expert elicitation as a means to reduce uncertainty and inform decision making [@Morgan_2014]. But experts can sometimes be mistaken, a fact implicitly recognized when patients request a "second opinion" from another doctor when facing uncertainty. Additionally, relying on only a small panel of experts for a problem as complex as malaria exposes one to bias, just as an overly concentrated investment portfolio exposes one to more risk. As Sir Francis Galton demonstrated in his famous study in which he showed that the crowds' aggregated estimates of cow's weight formed a quasi-normal distribution centered around the true weight [@Wallis2014] [@Galton1907Voxpopuli], averaging the perceptions of many can be more accurate than taking the opinion of any single expert, since the biases of diverse viewpoints can be complementary and symbiotic.

@@

The objective of this study is to gauge (expert) opinion about, estimate the likelihood of, and quantify the potential time frame to malaria eradication through a systemic survey of malaria research professionals from a wide array of academic disciplines. In doing so, we aim to help guide the optimal distribution of health resources by informing estimations of the expected value of malaria eradication efforts.

@@

1. Area of expertise.
2. Perceived probability (%) of malaria eradication in 10, 20, 30, 40, and 50 years.
3. Free choice perceived number of years until malaria eradication.
4. Ranking of the WHO's "health system building blocks" [@manya2017] in regards to attention needed in order to achieve eradication (the purpose of final point was to introduce a prescriptive element to the researchers' responses, ie, identify potential areas of consensus in regards to the path towards eradication).

The survey was intentionally as short as possible, so as to appeal to time-pressed participants. However, supplementary data on researchers is of value for the assessing of selection bias and determinants of perception. Accordingly, we estimated the likelihood of a participant being of one of two genders (male or female) based on first name, and the likelihood of a participant being of one of five ethnicities (White, Black, Hispanic, Asian, Other) based on the provided last name. For gender estimations we used data from the North Atlantic Population Project, and U.S. government [@Lincoln]; for ethnicity, Khanna et al's bayesian prediction algorithm for the estimation of racial categories [@wru2015] [@Imai_2016] was employed. The searching and retrieval of information from the PubMed database was carried out using the RISmed package [@Kovalchik2015].

@@

@@

\subsection*{Quantitative results}

@@

Respondents were qualitatively different from non-respondents. Importantly, the average number of total author-specific citations was `r round(mean(df$n_citations_total[df$responded]), digits = 2)` among respondents, but `r round(mean(df$n_citations_total[!df$responded]), digits = 2)` among non-respondents. This suggests a tendency for more senior or impactful researchers not to respond. When examining the number of average citations per article, the difference between respondents remained: `r round(mean(df$n_citations_average[df$responded]), digits = 2)` among respondents, and `r round(mean(df$n_citations_average[!df$responded]), digits = 2)` among non-respondents, highlighting the greater impact of non-respondents relative to respondents. Males responded at a greater rate (`r round(mean(df$responded[df$sex == 'male'], na.rm = T) * 100, digits = 2)`) than females (`r round(mean(df$responded[df$sex == 'female'], na.rm = T) * 100, digits = 2)`) and those with Hispanic last names responded at a greater rate than those with last names of other ethnicities (table 1).

@@

In order to de-bias sample selection, we employed Heckman's two-step correction method as well as simple binomial logistic regression model to estimate the likelihood of response as a function of sex and (binned) number of citations. To the extent that the two methods presented nearly identical value, we present here the results of the latter. Having estimated the odds of survey participation, we then used the inverse of the selection model's predictions as weights in a simple linear model to adjust our estimates. We run a separate weighted model to estimate the likelihood of eradication at 10, 20, 30, 40, and 50 years. Figure 3 shows both the aggregated perceived likelihood of eradication over time by number of publications (panel A) as well as the difference in the aggregate perceived likelihood of eradication for the entire sample before and after adjusting for sample selection (panel B).

@@

}
simple <- gather(simple, key, value, adjusted:unadjusted)
g2 <- ggplot(data = simple,
       aes(x = years,
           y = value,
           color = key,
           lty = key)) +
    geom_line(alpha = 0.8) +
  geom_point(alpha = 0.6) +
  theme_paper(base_size = 13) +
  labs(x = 'Years',
       y = 'Likelihood',
       subtitle = 'B') +
  scale_linetype_manual(name = '', values = 1:2) +
  scale_color_manual(name = '', values = c('blue', 'darkorange'))
x <- df %>%
  group_by(citations = bin_citations) %>%
  summarise(years_10 = mean(years_10_numeric, na.rm = TRUE),
            years_20 = mean(years_20_numeric, na.rm = TRUE),
            years_30 = mean(years_30_numeric, na.rm = TRUE),
            years_40 = mean(years_40_numeric, na.rm = TRUE),
            years_50 = mean(years_50_numeric, na.rm = TRUE),
            n = length(unique(email))) %>%
  ungroup %>%
  gather(key, value, years_10:years_50) %>%
  filter(!is.na(citations))

cols <- colorRampPalette(c('red', 'orange', 'blue') #brewer.pal(n = 9, 'Spectral')
                         )(length(unique(x$citations)))
g1 <- ggplot(data = x %>% filter(!is.na(citations)),
       aes(x = as.numeric(gsub('years_', '', key)),
           y = value,
           color = citations,
           group = citations)) +
  geom_point(alpha = 0.7) +
  geom_line(alpha = 0.6) +
  theme_paper(base_size = 13) +
  scale_color_manual(name = 'N° citations',
                     values = cols) +
  labs(x = 'Years',
       y = 'Likelihood',
       subtitle = 'A')
```

@@

2. *Systemic challenges*: comments pertaining to political, social, environmental or logistical issues related to eradication.

3. *Complexity*: comments which focus on the multi-dimensional components of eradication.

@@

Others echoed the sentiment, with many comments focusing on the need of strong surveillance and treatment delivery systems. Many commenters focused on other reasons for stagnating progress, such as "weak or failing health systems...due to political unwillingness or conflict".  Many noted that malaria is a "disease of poverty", with "social injustice" as the root cause. Some made the sequential argument that "eradication of poverty" must preceed disease eradication. Along the same lines, one commenter wrote:

@@

Another noted that the survey "left off the list the most important factor - economic development". Many echoed the sentiment, stating that "with no economic development, you cannot have eradication" and that poverty is the "cause" of malaria. Comments coded as "systemic" tended to be descriptive and more pessimistic than others.

@@

Comments questioning the utility of eradication as a concept or goal tended to be skeptical of its feasibility. Largely, they were prescriptive, advocating for a re-framing of the conversation so that the focus was not on an "arbitrary" goal like eradication, but rather on scaling up control and making region-specific progress.

@@

Approximately three-quarters of respondents believe that malaria will not be eradicated in the next 50 years; in other words, assuming our pool of respondents is of typical post-PhD age, most believe they will not live to see eradication. When adjusted for selection bias, the perceived likelihood of eradication 50 years from now remains similarly low, but estimates for shorter-term eradication are even lower.

If we take the pooled opinions of researchers, as revealed through this survey, as the best approximation of truth available, then the disconnect between public discourse on malaria eradication and private opinion is noteworthy This has important implications for spending in public health, since the attention and resources of funders, researchers and government agencies are directed to those areas where a result is expected.

@@

There are several reasons to be skeptical that the combined opinions of malaria researchers is a reliable indicator of truth. First, academic researchers are specialists - their narrow, field-specific view of eradication's feasibility is arguably less reliable than the more "global" views of large institutions. Though crowds have been found to be more "wise" than individuals in many cases, the application of a wisdom of crowds approach is not suitable to all classes of problems [@Mannes_2014]. Crowds can be susceptible to social biases [@Lorenz_2011] (although this survey's anonimity largely protects against this issue), and other biases may come in to play, especially given that our study was of a crowd of "specialists", rather than the population as a whole.

@@

Our survey attempted to quantify the likelihood of, and time frame to, malaria eradication by gauging the collective opinion of malaria researchers. Our survey shows that the opinions of malaria researchers appear to diverge from those of malaria programme financiers in regards to the feasibility and timeframe of eradication, with researchers expressing more scepticism than the institutions and individuals which finance and oversee global eradication efforts. The causes for scepticism are diverse, but common themes were the need for innovation, systemic challenges, and the complexity of the disease and its transmission.

@@

We have strived to present the results of our study without insinuation regarding the "true" feasibility and timeframe of eradication, as only time will tell whether the collective "wisdom" of researchers was worth adhering to or not. If the research community is unduly sceptical, efforts should be made to convince them that eradication is more feasible than they realize (so that resources are directed accordingly). If, on the other hand, collective wisdom of specialists is a better forecasting agent than the stated objectives of global health leaders and institutions, then it is important to recognize the sad truth that we are likely further from eradication than we would like to be. In this case, though the discourse on eradication can be motivating and inspiring, resources should be directed in a manner that reflects the intended outcome's likelihood of occurrence.
